---
title: "Using MUSE"
output: html_document
date: "2024-11-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R package, MUSE implements Joint Inverse Probability Weighted (JIPW) and Joint Augmented Inverse Probability Weighted (JAIPW) methods to address selection bias in non-probability samples, such as Electronic Health Records (EHRs). These methods leverage data integration techniques, incorporating either individual-level data or summary-level statistics from external sources, to improve the estimation of association parameters in binary disease risk models.

Selection bias poses significant challenges when working with EHR data, particularly when participants are recruited from multiple clinics or centers, each with distinct and potentially outcome-dependent selection mechanisms. Standard inverse-probability-weighted (IPW) methods, which rely on parametric selection models, often suffer from inaccuracies due to model misspecification in such settings.

The JAIPW method enhances robustness by integrating individual-level data from multiple cohorts with diverse selection mechanisms and supplementing it with data from an external probability sample. By incorporating a flexible auxiliary score model, JAIPW achieves double robustness, effectively mitigating biases caused by misspecified selection models and improving the reliability of analysis in complex, non-probability sampling scenarios.


Let \(D\) be a binary disease indicator, and let \(\boldsymbol{Z}\) denote a set of covariates. The primary disease model of interest in the target population is: 

\[
\text{logit}(P(D=1|\boldsymbol{Z})) = \theta_0 + \boldsymbol{\theta}_{\boldsymbol{Z}} \boldsymbol{Z}.
\]

We analyze data from \(K\) internal non-probability samples (cohorts) drawn from the same target population. Each cohort \(k\) (\(k = 1, 2, \ldots, K\)) is characterized by a binary selection indicator \(S_k\), which represents inclusion in the cohort. For each cohort \(k\), \(\boldsymbol{Z}_{1k}\) denotes the subset of covariates that appear exclusively in the disease model and do not directly influence the corresponding selection indicator \(S_k\). \(\boldsymbol{Z}_{2k}\) includes covariates shared by both the disease and selection models. Across all cohorts, the set of covariates \(\boldsymbol{Z}\) is the union of \(\boldsymbol{Z}_{1k}\) and \(\boldsymbol{Z}_{2k}\). Additionally, \(\boldsymbol{W}_k\) represents variables specific to the selection model for cohort \(k\), which can vary across cohorts. The probability of selection into cohort \(k\), given covariates, is modeled as \(P(S_k=1|\boldsymbol{X}_k) = \pi_k(\boldsymbol{X}_k)\), where \(\boldsymbol{X}_k\) includes \(D\), \(\boldsymbol{Z}_{2k}\), and \(\boldsymbol{W}_k\).

## Data Generation

The data is generated using the R script **"Data_gen_indi.R"**. For this example, we set the number of cohorts to \( K = 3 \), the population size to \( N = 50{,}000 \), and the dimension of the covariate vector to \(\text{dim}(\boldsymbol{Z}) = 3\). Different selection mechanisms are employed for the three cohorts to reflect cohort-specific variability. Additionally, we simulated individual-level external data to evaluate the methods JPL, JSR, and JAIPW, and generated marginal totals for JCL.


```{r,eval=FALSE}
source("Data_gen_indi.R")

theta ## actual theta_z
```

## Unweighted Logistic Method

At first, we implement an unweighted logistic method.

```{r}
source("MUSE_wrap.R") ## main function
```

```{r,eval=FALSE}
## Unweighted Logistic Method with variance estimate

res_uw=MUSE_func(K=K,N=N,Z_names=Z_names,
                 intdata_list=intdata_list,variance = TRUE)
res_uw

## set UW_CS=TRUE for cohort specific intercepts
```

## Inverse Probability Weighted Regression Using Individual level External data

## Pseudolikelihood (PL)

```{r,eval=FALSE}
## Approximate variance

res_pl= MUSE_func(K=K,N=N,Z_names=Z_names,intdata_list=intdata_list,IPW=TRUE,
              select_var_list=select_var_list,
              extdata=extdata,Weights_e = 1/extdata$Select_Weights,
              variance = TRUE,ipw_method ="PL")
res_pl

## Asymptotic variance

# res_pl= MUSE_func(K=K,N=N,Z_names=Z_names,intdata_list=intdata_list,IPW=TRUE,
#               select_var_list=select_var_list,
#               extdata=extdata,Weights_e = 1/extdata$Select_Weights,
#               variance = TRUE,ipw_method ="PL", type_var = "asy")

## Just change to ipw_method="SR" for Simplex Regression
```

## Inverse Probability Weighted Regression Using Marginal level external data

## Calibration (CL)

The data is generated using the R script **"Data_gen_marg.R"**

```{r,eval=FALSE}
source("Data_gen_marg.R")
```

```{r,eval=FALSE}

res_cl=MUSE_func(K=K,N=N,Z_names=Z_names,intdata_list=intdata_list,IPW=TRUE,
            select_var_list=select_var_list,
            marginals_list = margs,
            variance =TRUE,ipw_method = "CL")

```

## Joint Augmented Inverse Probability Weighted Method (JAIPW)

```{r,eval=FALSE}

res_aipw=MUSE_func(K=K,N=N,Z_names=Z_names,intdata_list=intdata_list,IPW=TRUE,
              select_var_list=select_var_list,aux_var_list=aux_var_list,
              extdata=extdata,Weights_e = 1/extdata$Select_Weights,
              variance = TRUE,ipw_method ="PL",AIPW=TRUE)

```


